\paperTitle

### 2014-05-17 doc/icml2014/background.tex
Updated intro figure, added exclusive views figure
some changes near beginning - bottlenecks, etc.

@@

We start by trying to reveal some information about the hidden variables
that will be used by subsequent sections.
Specifically, we review how
the tensor factorization method of \citet{anandkumar13tensor}
can be used to recover the conditional moments
$\mOpp{v}{i} = \Pr(x_v \mid h_i)$.
The key notion is that of a bottleneck:

\begin{definition}[Bottleneck]
  \label{def:bottleneck}
  A hidden variable $h_i$ is said to be a \emph{bottleneck} if there
  exists three observed variables (views), $x_{v_1}, x_{v_2}, x_{v_3}$
  that are conditionally independent given $h_i$,
  and each $\mOpp{v}{i} \eqdef \Pr(x_v \mid h_i) \in \Re^{d \times k}$ has full column rank $k$
  for each $v \in \{v_1, v_2, v_3\}$.
  %Let $\sV_{h_i}$ denote any set of views for $h_i$.
  We say that a subset of hidden variables $S \subseteq \bh$ is a bottleneck
  if every $h \in S$ is a bottleneck.
  We say that a graphical model $\sG$ is bottlenecked if all hidden variables
  are bottlenecks.
\end{definition}
For example, in \figureref{approach}, $x_1^a,x_1^b,x_2^a$ are views of the bottleneck $h_1$;
$x_2^a,x_2^b,x_1^b$ are view of the bottleneck $h_2$.  Therefore, the clique
$\{h_1,h_2\}$ is a bottleneck.  Note that views are allowed to overlap.

%Intuitively, we need to ensure that every marginal distribution
  %$\Pr(x_v)$ can be separated into $k$ different conditional distributions
  %$\Pr(x_v | h_i)$. 
The full rank assumption on the conditional moments $\mOpp{v}{i} = \Pr(x_v \mid h_i)$
ensures that all states of $h_i$ ``behave differently''.
In particular, the conditional distribution of one state cannot be
a mixture of that of other states.

\begin{figure}[t]
  \label{fig:three-view}
  \centering
  \input{figures/three-view.tikz}
  \caption{The canonical three-view mixture model can be estimated consistently
  using tensor factorization \citep{anandkumar13tensor}.}
\end{figure}

\citet{anandkumar12moments} provide an efficient tensor factorization algorithm
for estimating $\Pr(x_v \mid h_i)$:
  %using the method of moments to learn the parameters of a three-view
\begin{theorem}[Tensor factorization]
Let $h_i \in \bh$ be a hidden variable
with views $x_{v_1},x_{v_2},x_{v_3}$.
Then there exists an algorithm $\TensorFactorize$ that
returns consistent estimates $\mOpphat{v}{i}$ for each $v \in \{v_1,v_2,v_3\}$.
\end{theorem}

To simplify notation, consider the example in \figureref{three-view} where $h_1=1,v_1=1,v_2=2,v_3=3$.
The key idea is that the observed moments $\mO_{12}, \mO_{23}, \mO_{13}$ and
  $\mO_{123}$ admit a factorization in terms of parameters:
\begin{align*}
  \mO_{ij} &= \sum_h {\pi \oft 1}[h] {\mOppt{i}{1}}[h] \otimes {\mOppt{j}{1}}[h] \\ %\quad \forall i,j \in \{1,2,3\} \\
  \mO_{123} &= \sum_h {\pi \oft 1}[h] {\mOppt{1}{1}}[h] \otimes {\mOppt{2}{1}}[h] \otimes  {\mOppt{3}{1}}[h].
\end{align*}
The algorithm first computes a whitening matrix $W \in
  \Re^{d \times k}$ such that $W^\top \mO_{12} W = I_{k \times k}$,
  and uses $W$ to transform $\mO_{123}$ into a symmetric orthogonal tensor.  
Then a robust tensor power method is applied to extract the eigenvectors
of the whitened $\mO_{123}$; unwhitening yields the columns of $\mOpp{3}{1}$.
The other model parameters can be recovered with a few additional matrix operations.
%  we call $\TensorFactorize$, which transforms estimates of the observed moments 
%  into estimates of the parameters, $\pi\oft 1, \mOpp{1}{1},
%  \mOpp{2}{1},\mOpp{3}{1}$ (up to permutation of the columns), provided
%  the following assumption holds:

%It is easy to see that $\LearnMarginals$ has a polynomial sample complexity
%because it composes two parts that individually have polynomial sample
%complexity.
%From \citet{anandkumar13tensor} given the non-degeneracy assumptions
  %\assumptionref{non-degeneracy},
  We have that $\mOpphat{v}{i}$
  converges to $\mOpp{v}{i}$ at a rate of $n^{-\frac12}$ with a constant
  that depends polynomially on the $k$-th singular value of
  $\mOpp{v}{i}$.
% PL: we don't have a crisp theorem (because there's multiple paths), so leave it vague.
%Recovering the marginals $Z_\sC$ from the conditional moments via $\LearnClique$ is a linear operation and thus also has polynomial sample complexity. 
However, $\sigma_{k}(\mOpp{v}{i})$ can become extremely
small if $h_i$ and $x_v$ are connected via many intermediate hidden
variables.\footnote{To see this, suppose $h_1$ has a view $x_v$ via a chain:
$h_1 - h_2 \cdots - h_t - x_v$. In this example, if
$\sigma_k(\Pr(h_{i+1} \given h_{i})) = a_k$ for each $i = 1,
DEFINE
\cdots, t-1$, then $\sigma_k(\mOpp{v}{1}) = a_k^t \sigma_k(\mOpp{v}{t})$.}.

% No parameters yet
The tensor factorization method attacks the heart of the non-convexity
  in latent-variable models, providing some information about the hidden variables
  in the form of the conditional moments $\mOpp{v}{i} = \Pr(x_v \mid h_i)$.
  Note that $\TensorFactorize$ only looks at the conditional independence structure
  of the graphical model, not its parametrization.
  If $v$ is the single parent of $i$ (e.g., $\Pr(x_1^a \mid h_1)$ in \figureref{approach}),
  then this conditional moment is a parameter of the model.
  If not (e.g., $\Pr(x_2^a \mid h_1)$), then this is not a parameter.
  Furthermore, there are some parameters (e.g., $\Pr(h_4 \mid h_2, h_3)$) which are
  are still not recovered.
  In general, there is a gap between the conditional moments
  and the model parameters,
  something we will address in the next section.

>>>>

Finally, for clarity, we will derive our algorithms using exact moments of the true
distribution $p_{\theta^*}$.  In practice, we would use moments estimated from data $\sD$.

@@

We start by trying to reveal some information about the hidden variables
that will be used by subsequent sections.
Specifically, we review how
the tensor factorization method of \citet{anandkumar13tensor}
can be used to recover the conditional moments
$\mOpp{v}{i} \eqdef \Pr(x_v \mid h_i)$.
The key notion is that of a bottleneck:

\begin{definition}[Bottleneck]
  \label{def:bottleneck}
  A hidden variable $h_i$ is said to be a \emph{bottleneck} if (i) there
  exists three observed variables (views), $x_{v_1}, x_{v_2}, x_{v_3}$,
  that are conditionally independent given $h_i$,
  and (ii) each $\mOpp{v}{i} \eqdef \Pr(x_v \mid h_i) \in \Re^{d \times k}$ has full column rank $k$
  for each $v \in \{v_1, v_2, v_3\}$.
  %Let $\sV_{h_i}$ denote any set of views for $h_i$.
  We say that a subset of hidden variables $S \subseteq \bh$ is a bottleneck
  if every $h \in S$ is a bottleneck.
  We say that a graphical model $\sG$ is bottlenecked if all its hidden variables
  are bottlenecks.
\end{definition}
For example, in \figureref{approach}, $x_1^a,x_1^b,x_2^a$ are views of the bottleneck $h_1$,
and $x_2^a,x_2^b,x_1^b$ are views of the bottleneck $h_2$.  Therefore, the clique
$\{h_1,h_2\}$ is a bottleneck.  Note that views are allowed to overlap.

%Intuitively, we need to ensure that every marginal distribution
  %$\Pr(x_v)$ can be separated into $k$ different conditional distributions
  %$\Pr(x_v | h_i)$. 
The full rank assumption on the conditional moments $\mOpp{v}{i} = \Pr(x_v \mid h_i)$
ensures that all states of $h_i$ ``behave differently.''
In particular, the conditional distribution of one state cannot be
a mixture of that of other states.

%\begin{figure}[t]
%  \label{fig:three-view}
%  \centering
%  \input{figures/three-view.tikz}
%  \caption{The canonical three-view mixture model can be estimated consistently
%  using tensor factorization \citep{anandkumar13tensor}.}
%\end{figure}

\citet{anandkumar12moments} provide an efficient tensor factorization algorithm
for estimating $\Pr(x_v \mid h_i)$:
  %using the method of moments to learn the parameters of a three-view
\begin{theorem}[Tensor factorization]
Let $h_i \in \bh$ be a bottleneck with views $x_{v_1},x_{v_2},x_{v_3}$.
Then there exists an algorithm $\TensorFactorize$ that
returns consistent estimates of $\mOpp{v}{i}$ for each $v \in \{v_1,v_2,v_3\}$.
\end{theorem}

To simplify notation, consider the example in \figureref{three-view} where $h_1=1,v_1=1,v_2=2,v_3=3$.
The observed moments $\mO_{12}, \mO_{23}, \mO_{13}$ and $\mO_{123}$ can be factorized as follows:
\begin{align*}
  \mO_{vv'} &= \sum_h {\pi \oft 1}[h] {\mOppt{v}{1}}[h] \otimes {\mOppt{v'}{1}}[h] \\ %\quad \forall i,j \in \{1,2,3\} \\
  \mO_{123} &= \sum_h {\pi \oft 1}[h] {\mOppt{1}{1}}[h] \otimes {\mOppt{2}{1}}[h] \otimes  {\mOppt{3}{1}}[h].
\end{align*}
The $\TensorFactorize$ algorithm first computes a whitening matrix $W \in
  \Re^{d \times k}$ such that $W^\top \mO_{12} W = I_{k \times k}$,
  and uses $W$ to transform $\mO_{123}$ into a symmetric orthogonal tensor.  
Then a robust tensor power method is used to extract the eigenvectors
of the whitened $\mO_{123}$; unwhitening yields the columns of $\mOpp{3}{1}$
(up to permutation and scaling).
The other conditional moments can be recovered similarly.
%  we call $\TensorFactorize$, which transforms estimates of the observed moments 
%  into estimates of the parameters, $\pi\oft 1, \mOpp{1}{1},
%  \mOpp{2}{1},\mOpp{3}{1}$ (up to permutation of the columns), provided
%  the following assumption holds:

%It is easy to see that $\LearnMarginals$ has a polynomial sample complexity
%because it composes two parts that individually have polynomial sample
%complexity.
%From \citet{anandkumar13tensor} given the non-degeneracy assumptions
  %\assumptionref{non-degeneracy},
  The resulting estimate of %$\mOpphat{v}{i}$
  $\mOpp{v}{i}$ based on $n$ data points converges at a rate of $n^{-\frac12}$ with a constant
  that depends polynomially on $\sigma_k(\mOpp{v}{i})^{-1}$,
  the inverse of the $k$-th largest singular value of $\mOpp{v}{i}$.
% PL: we don't have a crisp theorem (because there's multiple paths), so leave it vague.
%Recovering the marginals $Z_\sC$ from the conditional moments via $\LearnClique$ is a linear operation and thus also has polynomial sample complexity. 
Note that $\sigma_{k}(\mOpp{v}{i})$ can become quite
small if $h_i$ and $x_v$ are connected via many intermediate hidden
variables.\footnote{To see this, suppose $h_1$ has a view $x_v$ via a chain:
$h_1 - h_2 \cdots - h_t - x_v$. In this example, if
$\sigma_k(\Pr(h_{i+1} \given h_{i})) = a_k$ for each $i = 1,
\cdots, t-1$, then $\sigma_k(\mOpp{v}{1}) = a_k^t \sigma_k(\mOpp{v}{t})$.}

% No parameters yet
The tensor factorization method attacks the heart of the non-convexity
  in latent-variable models, providing some information about the hidden variables
  in the form of the conditional moments $\mOpp{v}{i} = \Pr(x_v \mid h_i)$.
  Note that $\TensorFactorize$ only examines the conditional independence structure
  of the graphical model, not its parametrization.
  If $i$ is the single parent of $v$ (e.g., $\Pr(x_1^a \mid h_1)$ in \figureref{approach}),
  then this conditional moment is a parameter of the model,
  but this is in general not the case (e.g., $\Pr(x_2^a \mid h_1)$).
  Furthermore, there are other parameters (e.g., $\Pr(h_4 \mid h_2, h_3)$) which we do
  not have a handle on yet.
  In general, there is a gap between the conditional moments
  and the model parameters,
  something we will address in the next two sections.
